---
title: "NYPD Historical Shooting Data Analysis"
output: pdf_document
---
This notebook is a brief exploration of the NYPD Historical Shooting Data set.  It will focus on shooting related homocides and their patterns by precinct and in time.

# Loading Libraries
```{r, setup, results = "hide", message = FALSE, warning = FALSE}
# Load packages
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(MASS)
library(randomForest)
# Crate project 'data' folder if it does not exist
if (!dir.exists("data")) {
  dir.create("data")
}
```
# Loading Data
```{r, DATA-files, cache=TRUE}
# Define the URL
csv_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"

# Define file path
csv_file_path <- "data/rows.csv"

# Download CSV if it does not exist
if (!file.exists(csv_file_path)) {
  download.file(csv_url, csv_file_path)
}

# Load CSV
nydata <- read_csv(csv_file_path)
```
# Data Tidying Step

Here we will 'tidy' the data through formatting variables to the correct type, creating new variables, and dropping variables that won't be used in our analysis. We'll also check for missing variables but there may not be any missing data points in the columns we have selected for analysis.

```{r, tidy, dependson = c("DATA-files")}

startmissing <- sum(is.na(nydata))
print(paste("There are", startmissing, "missing data points in the starting data set."))
nydata.events <- nydata %>%

  # remove first set of unused columns
  dplyr::select(-JURISDICTION_CODE, -X_COORD_CD, -Y_COORD_CD, -Lon_Lat, -LOC_OF_OCCUR_DESC, -LOC_CLASSFCTN_DESC,   -LOCATION_DESC, -PERP_AGE_GROUP, -PERP_SEX, -PERP_RACE, -BORO, -VIC_RACE, -VIC_AGE_GROUP, -VIC_SEX) %>%



  # format and create date and time variables
  mutate(
          OCCUR_DATE = as.Date(as.character(OCCUR_DATE), format = "%m/%d/%Y"),
          OCCUR_TIME = hms(OCCUR_TIME),
          OCCUR_DATE_TIME = ymd_hm(paste(OCCUR_DATE, sprintf(
                                                            '%02d:%02d',
                                                            hour(OCCUR_TIME),
                                                            minute(OCCUR_TIME)
                                                            )
                                        )
          ),
          YEAR = year(OCCUR_DATE),
          MONTH = month(OCCUR_DATE),
          DAY = day(OCCUR_DATE),
          HOUR = hour(OCCUR_TIME),
          DAY_OF_WEEK = wday(OCCUR_DATE, label = TRUE),
          # The levels of the factor here are reversed so the weekend is on the outside of the radial graph
          DAY_OF_WEEK = factor(DAY_OF_WEEK, levels = c("Sun", "Sat", "Fri", "Thu", "Wed", "Tue", "Mon")),
  ) %>%

  # drop the handful of rows with no latitude or longitude
  drop_na(Latitude, Longitude) %>%

  arrange(OCCUR_DATE_TIME) %>%  # Make sure it's sorted by date-time
  distinct(INCIDENT_KEY, .keep_all = TRUE) %>%
  mutate(
    TIME_LAST_SHOOTING = as.numeric(difftime(OCCUR_DATE_TIME, lag(OCCUR_DATE_TIME, order_by = OCCUR_DATE_TIME, default = first(OCCUR_DATE_TIME)), units = "mins"))
  )

nydata.murders <- nydata.events %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  arrange(OCCUR_DATE_TIME) %>%
  distinct(INCIDENT_KEY, .keep_all = TRUE) %>%
  mutate(
    TIME_LAST_SHOOTING_MURDER = as.numeric(difftime(OCCUR_DATE_TIME, lag(OCCUR_DATE_TIME, order_by = OCCUR_DATE_TIME, default = first(OCCUR_DATE_TIME)), units = "mins"))
  )

# checking for missing data points in final dataset
totalmissing <- sum(is.na(nydata.murders))
print(paste("There are", totalmissing, "missing data points in the final data set."))

```
We've have no missing data points to worry about in the columns we have selected for anaylsis.  We can now begin to visually and analytically explore our data.


# Visualization #1
In this visualization we summarize the data to count the number of shooting deaths by hour and day of the week, and then create a polar plot to visualize the distribution.
```{r, polartime}
# Summarize the data to count the number of shootings by hour
shootings_by_hour <- nydata.events %>%
  group_by(DAY_OF_WEEK, HOUR) %>%
  summarize(n = n()) %>%
  ungroup()


# Create the polar plot with stacked bars
ggplot(shootings_by_hour, aes(x = as.factor(HOUR), y = n, fill = DAY_OF_WEEK)) +
  geom_bar(stat = "identity") +
  coord_polar("x") +
  scale_x_discrete(breaks = 0:23, labels = 0:23) +
  ggtitle("Number of Shootings by Hour and Day of the Week") +
  xlab("Hour of the Day") +
  ylab("Number of Shootings") +
  theme_minimal()


```
It appears that the majority of shooting related murders take place at night between 6p and 4a.

# Visualization #2
In this visualization we summarize the data to count the number of murders by precinct and create a histogram to visualize the distribution.
```{r, linegraph}
# Summarize the data to count the number of murders by precinct
murders_by_precinct <- nydata.murders %>%
  group_by(PRECINCT) %>%
  summarize(murder_count = n()) %>%
  ungroup()

# Create the histogram
ggplot(murders_by_precinct, aes(x = murder_count)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Murder Counts by Precinct") +
  xlab("Number of Murders") +
  ylab("Number of Precincts") +
  theme_minimal()
```


# Analysis and Modeling

Let's see if there is a correlation between day of the week and shooting murder counts using a random forest model.
```{r}
TOTAL_SHOOTINGS_PER_DAY <- nydata.murders %>%
        group_by(OCCUR_DATE, DAY_OF_WEEK) %>%
        summarize(murder_count = n()) %>%
        ungroup()

set.seed(777)
rf_model <- randomForest(murder_count ~ DAY_OF_WEEK, data = TOTAL_SHOOTINGS_PER_DAY, importance=TRUE)
print(rf_model)

predictions <- predict(rf_model, TOTAL_SHOOTINGS_PER_DAY)
observed <- TOTAL_SHOOTINGS_PER_DAY$murder_count
mae_value <- mean(abs(predictions - observed))

print(paste("Mean Absolute Error (MAE):", mae_value))
```

# A word on Bias

Potential sources of bias in this analysis of include underreporting of shootings and shootings that result in a murder. Additionally, there may be misclassified deaths where a murder is reported as a suicide or other administrative errors resulting in misreporting of data.

As this analysis spans many years differences in policing policy may result in changes in reporting for some years resulting in temporal bias.

One form of bias is confirmation bias where one tend to search for information that confirms one's preconceptions.  To mitigate this, actively seek out and consider data or viewpoints that challenge your existing beliefs.